---
title: "a bunch of data visualisations"
subtitle: "created from WHO/UNICEF JMP raw data collection"
author: "Lars Schöbitz"
format: 
  revealjs:
    smaller: true
    theme: slides.scss
    transition: fade
    slide-number: false
    footer: "https://github.com/larnsce/washr"
execute: 
  echo: false
editor: source
editor_options: 
  chunk_output_type: console
---

```{r}
#| include: false

library(tidyverse)
library(maps)
library(tmap) 
library(sf)
library(gt)

jmpraw <- read_rds(here::here("data/derived_data/2022-10-19_jmp_sanitation_raw_data.rds")) %>% 
    filter(year >= 2000, value >= 0) 


top10 <- function(data) {
    data  %>% 
        arrange(desc(n)) %>% 
        head(10)  %>% 
        mutate(country = countrycode::countrycode(iso3, origin = "iso3c", destination = "country.name")) %>% 
        select(country, n)
}


jmp_iso3_frequency <- jmpraw %>% 
    count(iso3)

jmp_iso3_user_interface_2015 <- jmpraw %>% 
    filter(year >= 2015) %>% 
    filter(san_service_chain == "user interface") %>% 
    count(iso3, san_service_chain)

jmp_iso3_emptying_2015 <- jmpraw %>% 
    filter(year >= 2015) %>% 
    filter(san_service_chain == "emptying") %>% 
    count(iso3, san_service_chain)

jmp_iso3_ww_treatment_2015 <- jmpraw %>% 
    filter(year >= 2015) %>% 
    filter(san_service_chain == "WW treatment") %>% 
    count(iso3, san_service_chain)

jmp_iso3_fs_treatment_2015 <- jmpraw %>% 
    filter(year >= 2015) %>% 
    filter(san_service_chain == "FS treatment") %>% 
    count(iso3, san_service_chain)


data("World")

world_moll = st_transform(World, crs = "+proj=moll")

tmap_mode("view")

tmap_options(basemaps = "OpenStreetMap")

tmap_mode("plot")

```

## R Data Package - Benefits

- Data accessible as a single table for any analysis tool
- Data can be imported to R using one command
- Public website with detailed documentation
    _ e.g. `washdata` R Package [https://katilingban.io/washdata/index.html](https://katilingban.io/washdata/index.html)

## R Data Package - Sanitation {.smaller}

- Data in long format (19,528 rows)
- 9 variables

```{r}

jmpraw %>% 
    relocate(iso3) %>%
    relocate(value, .after = san_service_chain) %>% 
    head(4) %>% 
    knitr::kable(digits = 1)

```

## R Data Package - New Variables {.smaller}

- `residence`: urban/rural/national
- `san_service_chain` (Sanitation Service Chain):

```{r}

jmpraw %>% 
    count(san_service_chain) %>% 
    knitr::kable()

```

## R Data Package - Use Cases {.smaller auto-animate="true"}

::: {.incremental}

1. Using JMP methods to reproduce estimates and apply different models (Linda Karani - MSc Data Science)

2. Writing an R Package with a function to produce estimates (and a function to produce service ladder plots) 


```{r}
#| eval: false
#| echo: true
#| fenced: true

```

:::

## R Data Package - Use Cases {.smaller auto-animate="true"}

1. Using JMP methods to reproduce estimates and apply different models (Linda Karani - MSc Data Science)

2. Writing an R Package with a function to produce estimates (and a function to produce service ladder plots) 

```{r}
#| eval: false
#| echo: true
#| fenced: true

estimate(iso3 = "AFG",           # default: all iso3 codes
         year = 2010:2030,       # Single year or range of years
         var_short = NULL,       # default: all variables (NULL)
         residence = "national") # default: national

```


## R Data Package - Use Cases {.smaller auto-animate="true"}

1. Using JMP methods to reproduce estimates and apply different models (Linda Karani - MSc Data Science)

2. Writing an R Package with a function to produce estimates (and a function to produce service ladder plots) 

```{r}
#| eval: false
#| echo: true
#| fenced: true

estimate(iso3 = "AFG",           # default: all iso3 codes
         year = 2010:2030,       # Single year or range of years
         var_short = NULL,       # default: all variables (NULL)
         residence = "national") # default: national

```

3. Great potential for unforeseen use cases enabled by making the data readily accessible (research, teaching, joining with other data, etc.)

## R Data Package - What's next?

- Submission of proposal for further development to [ORD (Open Research Data) Programme of ETH Domain](https://ethrat.ch/wp-content/uploads/2022/10/Application-guidelines_TrackC-Contribute-projects-3rd-call-1.pdf) (15k in-kind + 15k ETH Board), due: 12th December
- Submission of a proposal to Colorado WASH Symposium (focus on Linda's work), due: 25th November

::: {.aside}
https://ethrat.ch/wp-content/uploads/2022/10/Application-guidelines_TrackC-Contribute-projects-3rd-call-1.pdf
::: 

## Questions

1. How are country files updated on the server? 
    - How often? All together, or one by one as there is new data?

2. The current database for raw data shows 379 different sources. They are all abbreviated. Do you have a table where all these abbreviations are spelled out?

3. Ratio RS1

> "Other ratios used for 'basic' indicators (RW1, RS1) are calculated using simple averages" - JMP Methodology, March 2018


##

```{r}
jmpraw %>%
    count(type, sort = TRUE) %>% 
    knitr::kable()
    
```


## 

```{r}
jmpraw %>% 
    filter(!is.na(san_service_chain)) %>% 
    count(year, san_service_chain) %>% 
    group_by(year) %>% 
    mutate(
        prop = n / sum(n) * 100
    ) %>%  
    
    ggplot(aes(x = year, y = prop, fill = san_service_chain)) +
    geom_col() +
    labs(
        x = NULL,
        y = NULL, 
        title = "Proportion of data points along sanitation service chain",
        subtitle = "For all countries since 2000",
        fill = "Variable"
    ) +
    scale_fill_brewer(palette = "BrBG") +
    theme_minimal(base_size = 18) +
    theme(panel.grid.minor = element_blank())
```

##

```{r}
world_moll %>% 
    left_join(jmp_iso3_frequency, by = c("iso_a3" = "iso3")) %>% 
    filter(continent != "Antarctica") %>% 
    tm_shape() +
    tm_polygons(
        col = "n", 
        title = "No. of data points",
        #legend.hist = TRUE,
        #palette = "Oranges", 
        style = "fixed",
        contrast = c(0.1, 0.9),
        breaks = c(1, 100, 250, 500),
        #n = 4, 
        textNA = "No Data",
        colorNA = "grey80", 
        lwd = 0.5
    ) +
    tm_layout(
        main.title = "JMP raw data collection - Total number of data points since 2000",
        main.title.position = "center",
        title.position = c("left", "top"),
        #legend.outside = TRUE,
        scale = 1.5,
        #inner.margins = c(0, .02, .02, .02),
        #outer.margins = c(0, 0, 0, 0), frame = TRUE, 
    )

```

## 

```{r}
top10(jmp_iso3_frequency) %>% 
    knitr::kable()
```

##

```{r}
world_moll %>% 
    left_join(jmp_iso3_user_interface_2015, by = c("iso_a3" = "iso3")) %>% 
    filter(continent != "Antarctica") %>% 
    tm_shape() +
    tm_polygons(
        col = "n", 
        title = "No. of data points",
        #legend.hist = TRUE,
        #palette = "Oranges", 
        style = "fixed",
        contrast = c(0.1, 0.9),
        breaks = c(1, 11, 21, 31),
        #n = 4, 
        textNA = "No Data",
        colorNA = "grey80", 
        lwd = 0.5
    ) +
    tm_layout(
        main.title = "JMP raw data collection - Number of data points for 'user interface' since 2015",
        main.title.position = "center",
        title.position = c("left", "top"),
        #legend.outside = TRUE,
        scale = 1.5,
        #inner.margins = c(0, .02, .02, .02),
        #outer.margins = c(0, 0, 0, 0), frame = TRUE, 
    ) 

```

## 

```{r}
top10(jmp_iso3_user_interface_2015) %>% 
    knitr::kable()

```

##

```{r}

world_moll %>% 
    left_join(jmp_iso3_emptying_2015, by = c("iso_a3" = "iso3")) %>% 
    filter(continent != "Antarctica") %>% 
    tm_shape() +
    tm_polygons(
        col = "n", 
        title = "No. of data points",
        #legend.hist = TRUE,
        #palette = "Oranges", 
        style = "fixed",
        contrast = c(0.1, 1),
        breaks = c(1, 11, 21, 31),
        #n = 4, 
        textNA = "No Data",
        colorNA = "grey90", 
        lwd = 0.5
    ) +
    tm_layout(
        main.title = "JMP raw data collection - Number of data points for 'emptying' since 2015",
        main.title.position = "center",
        title.position = c("left", "top"),
        #legend.outside = TRUE,
        scale = 1.5,
        #inner.margins = c(0, .02, .02, .02),
        #outer.margins = c(0, 0, 0, 0), frame = TRUE, 
    ) 

```

##

```{r}
top10(jmp_iso3_emptying_2015) %>% 
    knitr::kable()
```

##

```{r}

world_moll %>% 
    left_join(jmp_iso3_ww_treatment_2015, by = c("iso_a3" = "iso3")) %>% 
    filter(continent != "Antarctica") %>% 
    tm_shape() +
    tm_polygons(
        col = "n", 
        title = "No. of data points",
        #legend.hist = TRUE,
        #palette = "Oranges", 
        style = "fixed",
        contrast = c(0.1, 1),
        breaks = c(1, 3, 6, 20),
        #n = 4, 
        textNA = "No Data",
        colorNA = "grey90", 
        lwd = 0.5
    ) +
    tm_layout(
        main.title = "JMP raw data collection - Number of data points for 'wastewater treatment' since 2015",
        main.title.position = "center",
        title.position = c("left", "top"),
        #legend.outside = TRUE,
        scale = 1.5,
        #inner.margins = c(0, .02, .02, .02),
        #outer.margins = c(0, 0, 0, 0), frame = TRUE, 
    )
```

##

```{r}
top10(jmp_iso3_ww_treatment_2015) %>% 
    knitr::kable()
```


##

```{r}
world_moll %>% 
    left_join(jmp_iso3_fs_treatment_2015, by = c("iso_a3" = "iso3")) %>% 
    filter(continent != "Antarctica") %>% 
    tm_shape() +
    tm_polygons(
        col = "n", 
        title = "No. of data points",
        #legend.hist = TRUE,
        #palette = "Oranges", 
        style = "fixed",
        contrast = c(0.1, 1),
        breaks = c(1, 3, 6, 20),
        #n = 4, 
        textNA = "No Data",
        colorNA = "grey90", 
        lwd = 0.5
    ) +
    tm_layout(
        main.title = "JMP raw data collection - Number of data points for 'faecal sludge treatment' since 2015",
        main.title.position = "center",
        title.position = c("left", "top"),
        #legend.outside = TRUE,
        scale = 1.5,
        #inner.margins = c(0, .02, .02, .02),
        #outer.margins = c(0, 0, 0, 0), frame = TRUE, 
    )
```

## 

```{r}
top10(jmp_iso3_fs_treatment_2015) %>% 
    knitr::kable()
```

## Thanks!

**Data source:** washdata.org

**Slides - source code:** https://github.com/larnsce/washr/blob/master/doc/slides/slides-jmp-raw-data-package.qmd

**Slides - PDF download:** https://github.com/larnsce/washr/raw/master/doc/slides/slides-jmp-raw-data-package.pdf

