---
title: "Data Vis"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(maps)
library(tmap)
library(ggalt)
library(spData)
library(tmap)   
library(sf)    

jmpraw <- read_rds(here::here("data/derived_data/2022-10-19_jmp_sanitation_raw_data.rds")) %>% 
    filter(year >= 2000, value >= 0) 


top10 <- function(data) {
    data  %>% 
        arrange(desc(n)) %>% 
        head(10)  %>% 
        mutate(country = countrycode::countrycode(iso3, origin = "iso3c", destination = "country.name")) %>% 
        select(country, n)
}


```

```{r}

# download indidivdual country files
# 
# name <- "GEO"
# 
# download.file(
#     url = paste0("https://washdata.org/data/country/", name, "/download"), 
#     destfile = here::here(paste0("data/raw_data/", name, ".xlsx")), mode = "wb"


```

# Explore data

Datasets included in the JMP database include:

- Censuses, which in principle collect basic data from all people living within a country. Censuses are always led by national statistical offices.
- Household surveys, which collect data from a subset of households. These may target national, rural, or urban populations, or more limited project or sub-national areas. An appropriate sample design is necessary for survey results to be representative, and surveys are often led by or reviewed and approved by national statistical organizations.
- Administrative data, may consist of information collected by government or non-government entities involved in the delivery or oversight of services. Examples include: water and sanitation inventories and databases, and reports of regulators.
- Other datasets may be available such as compilations by international or regional initiatives (e.g. IB-NET), studies conducted by research institutes, or technical advice received during country consultations.


```{r}

```


```{r}

## 5 NAs are countries without data

jmpraw %>% 
    count(type)

jmpraw %>% 
    count(type, source) %>% 
    arrange(desc(n))

jmpraw %>% 
    count(san_service_chain)

jmpraw %>% 
    ggplot(aes(x = year, fill = type)) +
    geom_bar()

jmpraw %>% 
    ggplot(aes(x = year, fill = san_service_chain)) +
    geom_bar()

```

```{r}

jmpraw %>% 
    filter(!is.na(san_service_chain)) %>% 
    group_by(san_service_chain) %>% 
    count() %>% 
    ggplot(aes(x = san_service_chain, y = n)) +
    geom_col() +
    labs(
        title = "Number of data sources along sanitation service chain",
        subtitle = "For 232 countries and since 2000"
    ) +
    annotate("text", x = 5, y = 1750, label = "388") +
    annotate("text", x = 6, y = 1750, label = "7") +
    annotate("text", x = 7, y = 1750, label = "58") +
    annotate("text", x = 8, y = 1750, label = "756") +
    theme_minimal()

```


```{r}

jmpraw %>% 
    filter(!is.na(san_service_chain)) %>% 
    count(san_service_chain) %>% 
    mutate(
        prop = n / sum(n) * 100
    ) %>% 
    
    ggplot(aes(x = san_service_chain, y = prop)) +
    geom_col() +
    labs(
        title = "Proportiong of data sources along sanitation service chain",
        subtitle = "For 232 countries and since 2015"
    )


```


```{r}

jmpraw %>% 
    filter(!is.na(san_service_chain)) %>% 
    count(year, san_service_chain) %>% 
    group_by(year) %>% 
    mutate(
        prop = n / sum(n) * 100
    ) %>%  
    
    ggplot(aes(x = year, y = prop, fill = san_service_chain)) +
    geom_col() +
    labs(
        title = "Proportiong of data sources along sanitation service chain",
        subtitle = "For 232 countries and since 2015"
    )


```



```{r}

jmp_iso3_frequency <- jmpraw %>% 
    count(iso3)

jmp_iso3_user_interface_2015 <- jmpraw %>% 
    filter(year >= 2015) %>% 
    filter(san_service_chain == "user interface") %>% 
    count(iso3, san_service_chain)

jmp_iso3_emptying_2015 <- jmpraw %>% 
    filter(year >= 2015) %>% 
    filter(san_service_chain == "emptying") %>% 
    count(iso3, san_service_chain)

jmp_iso3_ww_treatment_2015 <- jmpraw %>% 
    filter(year >= 2015) %>% 
    filter(san_service_chain == "WW treatment") %>% 
    count(iso3, san_service_chain)

jmp_iso3_fs_treatment_2015 <- jmpraw %>% 
    filter(year >= 2015) %>% 
    filter(san_service_chain == "FS treatment") %>% 
    count(iso3, san_service_chain)


```


# Maps

```{r}
## prepare world maps
## https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html


data("World")
world_moll = st_transform(World, crs = "+proj=moll")


iso3_country_names <- world_moll %>% 
    st_drop_geometry() %>% 
    as_tibble() %>% 
    select(iso3 = iso_a3, name)

countrycode::countrycode()


tmap_mode("view")
tmap_options(basemaps = "OpenStreetMap")
tmap_mode("plot")

```


## All data points since 2020

```{r}
world_moll %>% 
    left_join(jmp_iso3_frequency, by = c("iso_a3" = "iso3")) %>% 
    filter(continent != "Antarctica") %>% 
    tm_shape() +
    tm_polygons(
        col = "n", 
        title = "No. of data points",
        #legend.hist = TRUE,
        #palette = "Oranges", 
        style = "fixed",
        contrast = c(0.1, 0.9),
        breaks = c(1, 100, 250, 500),
        #n = 4, 
        textNA = "No Data",
        colorNA = "grey80", 
        lwd = 0.5
    ) +
    tm_layout(
        main.title = "JMP raw data collection - Total number of data points since 2000",
        main.title.position = "center",
        title.position = c("left", "top"),
        #legend.outside = TRUE,
        scale = 1.5,
        #inner.margins = c(0, .02, .02, .02),
        #outer.margins = c(0, 0, 0, 0), frame = TRUE, 
    )

top10(jmp_iso3_frequency)
```

## Data points for user interface since 2015

```{r}

world_moll %>% 
    left_join(jmp_iso3_user_interface_2015, by = c("iso_a3" = "iso3")) %>% 
    filter(continent != "Antarctica") %>% 
    tm_shape() +
    tm_polygons(
        col = "n", 
        title = "No. of data points",
        #legend.hist = TRUE,
        #palette = "Oranges", 
        style = "fixed",
        contrast = c(0.1, 0.9),
        breaks = c(1, 11, 21, 31),
        #n = 4, 
        textNA = "No Data",
        colorNA = "grey80", 
        lwd = 0.5
    ) +
    tm_layout(
        main.title = "JMP raw data collection - Number of data points for 'user interface' since 2015",
        main.title.position = "center",
        title.position = c("left", "top"),
        #legend.outside = TRUE,
        scale = 1.5,
        #inner.margins = c(0, .02, .02, .02),
        #outer.margins = c(0, 0, 0, 0), frame = TRUE, 
    ) 

top10(jmp_iso3_user_interface_2015)

```

## Data points for emptying since 2015

```{r}

world_moll %>% 
    left_join(jmp_iso3_emptying_2015, by = c("iso_a3" = "iso3")) %>% 
    filter(continent != "Antarctica") %>% 
    tm_shape() +
    tm_polygons(
        col = "n", 
        title = "No. of data points",
        #legend.hist = TRUE,
        #palette = "Oranges", 
        style = "fixed",
        contrast = c(0.1, 1),
        breaks = c(1, 11, 21, 31),
        #n = 4, 
        textNA = "No Data",
        colorNA = "grey90", 
        lwd = 0.5
    ) +
    tm_layout(
        main.title = "JMP raw data collection - Number of data points for 'emptying' since 2015",
        main.title.position = "center",
        title.position = c("left", "top"),
        #legend.outside = TRUE,
        scale = 1.5,
        #inner.margins = c(0, .02, .02, .02),
        #outer.margins = c(0, 0, 0, 0), frame = TRUE, 
    ) 

top10(jmp_iso3_emptying_2015)

```

## Data points for wastewater treatmnent since 

```{r}

world_moll %>% 
    left_join(jmp_iso3_ww_treatment_2015, by = c("iso_a3" = "iso3")) %>% 
    filter(continent != "Antarctica") %>% 
    tm_shape() +
    tm_polygons(
        col = "n", 
        title = "No. of data points",
        #legend.hist = TRUE,
        #palette = "Oranges", 
        style = "fixed",
        contrast = c(0.1, 1),
        breaks = c(1, 3, 6, 20),
        #n = 4, 
        textNA = "No Data",
        colorNA = "grey90", 
        lwd = 0.5
    ) +
    tm_layout(
        main.title = "JMP raw data collection - Number of data points for 'wastewater treatment' since 2015",
        main.title.position = "center",
        title.position = c("left", "top"),
        #legend.outside = TRUE,
        scale = 1.5,
        #inner.margins = c(0, .02, .02, .02),
        #outer.margins = c(0, 0, 0, 0), frame = TRUE, 
    ) 

top10(jmp_iso3_ww_treatment_2015)

```


```{r}

world_moll %>% 
    left_join(jmp_iso3_fs_treatment_2015, by = c("iso_a3" = "iso3")) %>% 
    filter(continent != "Antarctica") %>% 
    tm_shape() +
    tm_polygons(
        col = "n", 
        title = "No. of data points",
        #legend.hist = TRUE,
        #palette = "Oranges", 
        style = "fixed",
        contrast = c(0.1, 1),
        breaks = c(1, 3, 6, 20),
        #n = 4, 
        textNA = "No Data",
        colorNA = "grey90", 
        lwd = 0.5
    ) +
    tm_layout(
        main.title = "JMP raw data collection - Number of data points for 'faecal sludge' since 2015",
        main.title.position = "center",
        title.position = c("left", "top"),
        #legend.outside = TRUE,
        scale = 1.5,
        #inner.margins = c(0, .02, .02, .02),
        #outer.margins = c(0, 0, 0, 0), frame = TRUE, 
    ) 

top10(jmp_iso3_fs_treatment_2015)

jmpraw %>% 
    filter(iso3 == "BTN") %>% 
    filter(san_service_chain == "FS treatment")

```

```{r}
world_ww_treatment <- World %>% 
left_join(ww_treatment, by = c("iso_a3" = "iso3")) %>% 
filter(continent != "Antarctica") 

world_fs_treatment <- World %>% 
left_join(fs_treatment, by = c("iso_a3" = "iso3")) %>% 
filter(continent != "Antarctica")

tm_shape(world_ww_treatment) +
tm_polygons(col = "n", title = "Data points")

tm_shape(world_fs_treatment) +
tm_polygons(col = "n", title = "Data points")

```



